\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events

### 2023-09-19 paper/main.tex
expanded discussion section, added descriptions of replication experiment to methods section, added a note on odds ratios to methods

We asked participants to watch sequences of movie segments from a
character-driven television drama and then either retrodict what had happened
prior to a just-watched segment, predict what would happen next, or recall what
they had just watched. We found that participants tended to more accurately and
more readily retrodict the unobserved past than predict the unobserved future.
We traced this temporal asymmetry to (a) characters' tendencies to refer to
past events more than future events in their ongoing conversations, and (b)
associations between temporally proximal events (Fig.~\ref{fig:discussion}).
Essentially, associations between temporally proximal events serve to enhance
asymmetries in inferences driven by conversational references (light orange and
blue bars in Fig.~\ref{fig:discussion}). Our findings show that other peoples'
psychological arrows of time can affect external observers' inferences about
the unobserved past and future.

@@

@@

A total of 36 participants (25 female, mean age 21.47 years, range 19--50
years) were recruited from the Dartmouth College community. All participants
had self-reported normal or corrected-to-normal vision, hearing, and memory,
and had not watched any episodes of \textit{Why Women Kill} before the
experiment. Participants gave written consent to enroll in the study under a
protocol approved by the Committee for the Protection of Human Subjects at
Dartmouth College. Participants received course credit or monetary compensation
for their time. Two participants completed only the first half of the study and
one participant’s data from the second half of their testing session was lost
due to a technical error. All available data were used in the analyses.

@@

The stimulus used in the study were segments of the CBS television series
\textit{Why Women Kill} Season 1. The TV series contained three distinct
storylines depicting three women’s marital relationships. The three storylines,
which took place in the 1960s, 1980s, and 2019, were shown in an interleaved
fashion in the original episodes. The first 11 segments from the 1960s and
1980s storylines, across the first and second episodes, were used in our study.
Segments were divided based on major scene cuts, which primarily corresponded
to storyline shifts in the original episodes. The mean length of the segments
was 2.05 min (range 0.97--3.87 min). We chose this TV series based on its
strictly linear storytelling (within each storyline) and its realistic settings
where most events depicted everyday life. The plots were focused on the main
characters (Beth in storyline 1 and Simone in storyline 2), who were present in
all the segments in the corresponding storylines.

@@

Our experimental paradigm was divided across two testing sessions. In each
session, participants performed a sequence of tasks on segments from one
storyline (Fig.~\ref{fig:method}). For each storyline, there were four
different task sequences: two forward chronological order sequences and two
backward chronological order sequences. Participants completed one task
sequence in forward chronological order for one storyline, and one in backward
chronological order for the other storyline. The order of the two sessions
(forward chronological order sequence first or backward chronological order
sequence first), and the pairing of task sequences with storylines, were
counterbalanced across participants.

@@

@@

Events in the first 11 segments of the two storylines were identified by the
first author (X.X.), corresponding to major plot points (total: 117; mean: 5.32
per segment; range 3--9). Additionally, 74 offscreen events were identified. Of
these 74 offscreen events, 43 events were identified from references in
conversations during onscreen events. Another 16 events were identified based
on characters’ implied movements and travels. For example, if in segment 1
character A was in place A and in segment 2 she was in place B, then the
transit from place A to B for character A would be identified as an offscreen
event. The remaining 15 offscreen events were identified based on logical
inferences. For example, if a photograph was shown in an onscreen event (but
not the act of the photograph being taken), then the action that someone took
the photograph would be identified as an offscreen event. Offscreen events
always occurred between two contiguous segments, or before the first segment.
The purpose of identifying offscreen events was to match participants’
responses to video events; thus our identification of these offscreen events
was not intended to be exhaustive.

@@

Two coders (X.X. and Z.Z.) identified character dialogues in the narrative that
referred to past events or future (onscreen or offscreen) events. Only
references to events that occurred in a different segment were included in this
tagging procedure. For each reference, the source (referring) segment and the
referred event number were recorded. A total of 82 references were identified.
Of these, 30 referred to onscreen events and 52 referred to offscreen events.
For these referenced events, their corresponding summary events or partial
events were also labelled as referenced. In instances where the coders
disagreed about a given tag, disagreements were resolved through discussions
between the two coders. In our analyses, each storyline event was coded
according to whether or not it had been referenced in the segment(s) that the
participant had viewed thus far in the experiment.

@@

>>>>

We asked participants in our main experiment to watch sequences of movie
segments from a character-driven television drama and then either retrodict
what had happened prior to a just-watched segment, predict what would happen
next, or recall what they had just watched. We found that participants tended
to more accurately and more readily retrodict the unobserved past than predict
the unobserved future. We traced this temporal asymmetry to (a) characters'
tendencies to refer to past events more than future events in their ongoing
conversations, and (b) associations between temporally proximal events
(Fig.~\ref{fig:discussion}). Essentially, associations between temporally
proximal events serve to enhance asymmetries in inferences driven by
conversational references (light orange and blue bars in
Fig.~\ref{fig:discussion}). Our findings show that other peoples' psychological
arrows of time can affect external observers' inferences about the unobserved
past and future. We confirmed our main behavioral findings in a pre-registered
replication study. We also carried out a meta analysis of tens of millions of
utterances from television shows, movies, novels, and natural spoken and
written conversations. We found that people's tendencies to refer more often to
the past than the future appears to be a widespread characteristic of human
conversation.

@@

Beyond forming inferences about unobserved past and future events, our work
also relates to prior studies of how people perceive time~\citep{BlocGrub14,
Howa18, Eagl08, IvrySchl08, Wear16}, and how we ``move'' through time in our
memories of our past experiences~\citep{Mann21a, MannEtal11, HowaEtal12,
MannEtal07, ShanHowa12, Kaha96, PolyKaha08, SchaTulv94} or in our imagined
(past or future) experiences~\citep{Scha12, JossTone20, SchaEtal98,
MomeHowa18}. For example, a well-studied phenomenon in the episodic memory
literature concerns how remembering a given event cues our memories of other
events that we experienced nearby in time~\citep[i.e., the \textit{contiguity
effect};][]{Kaha96}. Across a large number of studies there appears to be a
nearly universal tendency for people to move \textit{forwards} in time in their
memories, whereby recalling an ``event'' (e.g., a word on a previously studied
list) is about twice as likely to be followed by recalling the event that
immediately followed as compared with the event immediately preceding the
just-recalled event~\citep{HealKaha14a}. Superficially our current study
appears to report the \textit{opposite} pattern, whereby participants display a
\textit{backwards} temporal bias. However, the two sets of findings may be
reconciled when one considers the frame of reference~\citep[and current mental
context; e.g.,][]{HowaKaha02a} of the participant at the moment they make their
response. In our study, participants observe an event in the present, and they
make guesses about what happened in the unobserved past or future, relative to
the just-observed event. (Our findings imply that participants are more facile
at moving backwards in time than forwards in time, relative to ``now.'') In
contrast, the classic contiguity effect in episodic memory studies refers to
how people move through time relative to a just \textit{remembered} event. The
forward asymmetry in the contiguity effect follows from the notion that the
moment of remembering has greater contextual overlap with events \textit{after}
the remembered event from the past than events that happened before
it~\citep[for review also see ][]{MannEtal15, Mann20}.

@@

\paragraph*{Main experiment.} A total of 36 participants (25 female, mean age
21.47 years, range 19--50 years) were recruited from the Dartmouth College
community for our main experiment. All participants had self-reported normal or
corrected-to-normal vision, hearing, and memory, and had not watched any
episodes of \textit{Why Women Kill} before the experiment. Participants gave
written consent to enroll in the study under a protocol approved by the
Committee for the Protection of Human Subjects at Dartmouth College.
Participants received course credit or monetary compensation for their time.
Two participants completed only the first half of the study and one
participant’s data from the second half of their testing session was lost due
to a technical error. All available data were used in the analyses.

\paragraph*{Replication experiment.} A total of XXX participants (XXX female,
mean age XXX years, range XXX--XXX years) were recruited from the Dartmouth
College community for our pre-registered replication experiment. All
participants had self-reported normal or corrected-to-normal vision, hearing,
and memory, and had not watched any episodes of \textit{The Chair} before the
experiment. Participants gave written consent to enroll in the study under a
protocol approved by the Committee for the Protection of Human Subjects at
Dartmouth College. Participants received monetary compensation for their time.
All available data were used in the analyses.

@@

\paragraph{Main experiment.}
The stimuli used in our main experiment were segments of the CBS television series
\textit{Why Women Kill} Season 1. The TV series contained three distinct
storylines depicting three women’s marital relationships. The three storylines,
which took place in the 1960s, 1980s, and 2019, were shown in an interleaved
fashion in the original episodes. The first 11 segments from the 1960s and
1980s storylines, across the first and second episodes, were used in our study.
Segments were divided based on major scene cuts, which primarily corresponded
to storyline shifts in the original episodes. The mean length of the segments
was 2.05 min (range 0.97--3.87 min). We chose this TV series based on its
strictly linear storytelling (within each storyline) and its realistic settings
where most events depicted everyday life. The plots were focused on the main
characters (Beth in storyline 1 and Simone in storyline 2), who were present in
all the segments in the corresponding storylines.

\paragraph{Replication experiment.} The stimuli used in our replication
experiment were segments of the first episodie of the Netflix television show
\textit{The Chair}, Season 1. \textbf{JRM NOTE: Describe the show, like you did
for Why Women Kill.} The mean length of the segments was XXX min (range
XXX--XXX min). As for the stimulus we used in our main experiment, we chose
this stimulus for our replication experiment for its linear storytelling
(again, within each storyline) and its realistic depections of everyday events.
\textit{JRM NOTE: The plots were focused on... (fill in something analogous to
what you wrote for the main experiment stimulus...)}

@@

\paragraph{Main experiment.}
Our experimental paradigm was divided across two testing sessions. In each
session, participants performed a sequence of tasks on segments from one
storyline (Fig.~\ref{fig:method}). For each storyline, there were four
different task sequences: two forward chronological order sequences and two
backward chronological order sequences. Participants completed one task
sequence in forward chronological order for one storyline, and one in backward
chronological order for the other storyline. The order of the two sessions
(forward chronological order sequence first or backward chronological order
sequence first), and the pairing of task sequences with storylines, were
counterbalanced across participants.

@@

\paragraph{Replication experiment.} \textbf{JRM NOTE: briefly describe
replication experiment methods, referring to Fig.~\MethodsReplExp. Since the
methods will have been similar for the replication study, just highlight the
differences rather than re-describing everything.}

@@

\paragraph{Main experiment.}
Events in the first 11 segments of the two storylines were identified by the
first author (X.X.), corresponding to major plot points (total: 117; mean: 5.32
per segment; range 3--9). Additionally, 74 offscreen events were identified. Of
these 74 offscreen events, 43 events were identified from references in
conversations during onscreen events. Another 16 events were identified based
on characters’ implied movements and travels. For example, if in segment 1
character A was in place A and in segment 2 she was in place B, then the
transit from place A to B for character A would be identified as an offscreen
event. The remaining 15 offscreen events were identified based on logical
inferences. For example, if a photograph was shown in an onscreen event (but
not the act of the photograph being taken), then the action that someone took
the photograph would be identified as an offscreen event. Offscreen events
always occurred between two contiguous segments, or before the first segment.
The purpose of identifying offscreen events was to match participants’
responses to video events; thus our identification of these offscreen events
was not intended to be exhaustive.

\paragraph{Replication experiment.} Events in the first XXX segments of the two
storylines were identified by the first author (X.X.), corresponding to major
plot points (total: XXX; mean: XXX per segment; range XXX--XXX). Additionally,
XXX offscreen events were identified. Of these XXX offscreen events, XXX events
were identified from references in conversations during onscreen events.
Another XXX events were identified based on characters’ implied movements and
travels. The remaining XXX offscreen events were identified based on logical
inferences.

@@

Two coders (main experiment: X.X. and Z.Z.; replication experiment: X.X. and
X.Z.) identified character dialogues in the narrative that referred to past
events or future (onscreen or offscreen) events. Only references to events that
occurred in a different segment were included in this tagging procedure. For
each reference, the source (referring) segment and the referred event number
were recorded. A total of 82 references were identified in the main experiment
stimulus, and XXX were identified in the replication experiment stimulus. Of
these references in the main experiment, 30 referred to onscreen events and 52
referred to offscreen events. In the replication experiment, XXX referred to
onscreen events and XXX referred to offscreen events. For these referenced
events, their corresponding summary events or partial events were also labelled
as referenced. In instances where the coders disagreed about a given tag,
disagreements were resolved through discussions between the two coders. In our
analyses, each storyline event was coded according to whether or not it had
been referenced in the segment(s) that the participant had viewed thus far in
the experiment.

@@

Several of our analyses entailed comparing the relative hit rates or
probabilities of two different conditions or outcomes. We used the
\texttt{emmeans} package to compute the odds ratios given the generalized
linear mixed models we fit for the given analysis. These odds ratios reflect
the chances (``odds'') of a particular outcome (e.g., making a response about a
particular event) given a scenario (e.g., the event occurred \textit{prior} to
the just-watched segment) compared with the chances of the outcome occurring in
the alternative scenario (e.g., the event occurred \textit{after} the
just-watched segment).
